### 2023-04-25 paper.tex
rm claim that our matrix works with seurat & rm a repetition

Lukas P.~M. Kremer\textsuperscript{1,2,*}, Martina Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
\vspace{1.5ex}

@@

In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data. We suggest improvements to current approaches, and demonstrate their value in benchmarks, using two real-world datasets. Furthermore, we discuss how to perform comparative analyses. Finally, we present ``scbs'', a comprehensive software toolkit to perform scBS data analysis, using the described methods. 
\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.

The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering. For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.

@@

To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
FDR estimation via permutation allows us to report statistical significance of each DMR.
% maybe here discuss pseudoreplication issues?
Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting. This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} \todo{bridges a gap in the chain of existing tools} that so far hindered practitioners in their data analysis.

@@

In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and \todo{bridges a gap in existing tool chains}.

>>>>

Lukas P.~M. Kremer\textsuperscript{1,2,*}, Martina M. Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex] 
\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
\vspace{1.5ex}

@@

In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data. We suggest improvements to current approaches, and demonstrate their value in benchmarks, using two real-world datasets. Furthermore, we discuss how to perform comparative analyses. Finally, we present ``\texttt{scbs}'', a comprehensive software toolkit to perform scBS data analysis, using the described methods. 
\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both, the simple percentage of methylated CpG sites, as well as our proposed methylation measure, the shrunken mean of the residuals, which is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any set of genomic features of interest, such as promoters, enhancers or transcription factor binding sites, obtaining one profile plot per cell.
% This is not really true cause you still need to deal with missing values and all the default transformations don't make sense when you're not dealing with count data: "The methylation matrix produced by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or ScanPy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering. For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.
The methylation matrix produced by the \texttt{matrix} function can then serve as input for established methods used for the exploration of single-cell data, such as dimensionality reduction and cell clustering.

@@

To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
FDR estimation via permutation allows us to report statistical significance of each DMR.
Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms.
The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting. This allows the user to use well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses, etc.
By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.

@@

In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these.
